Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-09-07 paper-correlations.tex
ААААА

\begin{abstract}
Correlation functions are becoming one of the major tools for quantification of
structural information that is usually represented as material's 2D or 3D image.
In this paper we introduce \code{CorrelationFunctions.jl} open-source package
developed in Julia and capable of computing all classical correlation functions
based on imaging input data. Images include both binary and multi-phase
representations.  Our code is capable of evaluating two-point probability $S_2$,
cluster $C_2$, lineal-path $L_2$, surface-surface $F_{ss}$, surface-void
$F_{sv}$, pore-size $P$ and chord-length $p$ functions on both CPU and GPU
architectures.  There possible, we presented two types of computations: full
correlation map (correlations of each point with other pints on the image, that
also allows obtaining ensemble averaged CF) and directional correlation
functions (currently in major orthogonal and diagonal directions).  Such an
implementation allowed for the first time to assemble a completely free solution
to evaluate correlation functions under any operating system with well
documented application programming interface (API). Our package includes
automatic tests against analytical solutions that are described in the paper. We
measured wall times for all CPU and GPU implementations and as a rule of thumb
full correlation maps on GPU are faster than other methods. However, full maps
require more RAM and, thus, are limited to available RAM resources. On the other
hand, directional CFs are memory efficient and can be evaluated for huge
datasets -- this way they are the first candidates for structural data
compression of feature extraction. The package itself is available through Julia
package ecosystem and on GitHub
(https://github.com/fatimp/CorrelationFunctions.jl), the latter source also
contains documentation and additional helpful resources such as tutorials.  We
believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} will significantly facilitate the usage of
correlation functions in numerous areas of structural description and research,
as well as in machine learning applications. We also present some example as
applied to ceramic, soil composite and oil-bearing rock samples based on their
3D X-ray tomography and 2D scanning electron microscope images. Finally, we
conclude our paper with discussion of possible ways to further improve presented
computational framework.
\end{abstract}

@@

Correlation functions as invaluable universal descriptors of structure are
utilized in a multitude of scientific disciplines: material sciences
\cite{hasanabadi20163d,Havelka,feng2018reconstruction,chen2022quantifying}, rock
physics \cite{ledesma2018effect}, soil physics and hydrology
\cite{PLoS_ONE,KarsaninaEJSS}, cosmology and food engineering
\cite{TakadaJain,Derossi2019}, biology \cite{veatch2012correlation} and many
others. Computed from 2D and 3D images, CFs can be applied to:
\begin{enumerate}
  \item to characterize the morphology and representativeness via correlation
    lengths \cite{vcapek2009stochastic,thovert2011grain,tensorPRE};
  \item to perform stochastic reconstructions from experimentally measured CFs
    or 2D to 3D reconstructions based on CFs that can be extracted from less
    dimensions or through penetrable sphere model
    \cite{Adler_recon,Y-T,tahmasebiPRL,hasanabadi20163d,karsaninaPRL};
  \item compare different structures to each other, including verification of
    stochastic reconstructions \cite{vcapek2009stochastic,tahmasebiPRL,EPL2};
  \item compress structural information in the form of raw or parameterized CFs
    with the possibility to recover the structure using stochastic
    reconstruction \cite{SciRep1,Havelka,KarsaninaEJSS};
  \item describe structural dynamics under different boundary conditions
    \cite{chen2015dynamic,xu2022correlation,fomin2023soil};
  \item extract structural features for deep learning
    \cite{pilania2017multi,kamrava2020linking,roding2020predicting,KarsaninaEJSS};
  \item fuse multi-scale images and structural information as obtained by
    different methods into a single digital model \cite{Geoderma2018};
  \item quantify spatial heterogeneity \cite{zhang2000pore,REVpaper}.
\end{enumerate}

@@

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width $7$ and
all its coefficients with exception of the central coefficient are inversely
proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_k &= S \left\{
    \begin{array}{ll}
      -\sum\limits_{\substack{l \\ l \ne 0}} E_l & \quad k = 0 \\
      1 / \rho(k, 0) & \quad \text{otherwise}
    \end{array}
    \right. \\
    k &\in \{-3, -2, \dots, 3\}^D
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $S$ equals to $30.45849$ in 2D and $172.96232$ in 3D case. The parameters
of the filter were adjusted based on comparison against exact computations of
surface functions on smooth-boundary sets \cite{postcov23}.

@@

The interface between phases must not change rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to be with an apropriate resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. We have previously
shown that criterion $C_{0.5} > 0.97$ holds then the image is suitable for
robust $F_{ss}$ and $F_{sv}$ evaluation \cite{samarin2023robust}.  Note that
function \code{lowfreq\_energy\_ratio} that computes $C_{0.5}$ criterion for any
input binary image is the part of \code{CorrelationFunctions.jl} package.

@@

All measurements of execution time are presented on \cref{fig:timings}. It is
immediately apparent that GPU versions (where available) are much faster than
CPU versions and limited only by amount of video memory. Functions in module
\code{Directional} (which utilize per-slice directional approach) are the
slowest among other implementations, but require less memory. Depending on CF,
our video card with 24 GBs of memory could handle a cube of $\sim 300^3$
voxels, whilst the workstation with 512-1024 GBs of RAM can handle virtually any
XCT-obtained image volume with any correlation length.

@@

\section{Summary}
\label{sec:summary}
In this paper we introduced \code{CorrelationFunctions.jl} open-source package
developed in Julia and capable of computing all classical correlation functions
based on 2D and 3D input images. Our code is capable of evaluating two-point
probability, cluster, lineal-path, surface-surface, surface-void, pore-size and
chord-length functions on both CPU and GPU architectures. There possible, we
presented two types of computations: full correlation map (correlations of each
point with other pints on the image, that also allows obtaining ensemble
averaged CF) and directional correlation functions (currently in major
orthogonal and diagonal directions). Such an implementation allowed for the
first time to assemble a completely free solution to evaluate correlation
functions under any operating system with well documented API. Our package
includes automatic tests against analytical solutions that are described in this
paper. We measured wall times for all CPU and GPU implementations and as a rule
of thumb full correlation maps are faster than directional scans because
calculation of correlation maps applies fast Fourier transform to the whole
input array while, when calculating scans, FFT needs to be applied to each scan
separately. Calculation on GPU shows performance increase when compared to CPU,
as expected.  However, full maps require more RAM and, thus, are limited to
available RAM resources. On the other hand, directional CFs are memory efficient
and can be evaluated for huge datasets -- this way they are the first candidates
for structural data compression of feature extraction. The package itself is
available through Julia package ecosystem and on GitHub
(https://github.com/fatimp/CorrelationFunctions.jl), the latter source also
contains documentation and additional helpful resources such as tutorials.  We
believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} will significantly facilitate the usage of
correlation functions in numerous areas of structural description and research,
as well as in machine learning applications.

>>>>

\begin{abstract}
Correlation functions are becoming one of the major tools for quantification of
structural information that is usually represented as material's 2D or 3D image.
In this paper we introduce \code{CorrelationFunctions.jl} open-source package
developed in Julia and capable of computing all classical correlation functions
based on imaging input data. Images include both binary and multi-phase
representations.  Our code is capable of evaluating two-point probability $S_2$,
cluster $C_2$, lineal-path $L_2$, surface-surface $F_{ss}$, surface-void
$F_{sv}$, pore-size $P$ and chord-length $p$ functions on both CPU and GPU
architectures.  Where possible, we presented two types of computations: full
correlation map (correlations of each point with other points on the image, that
also allows obtaining ensemble averaged CF) and directional correlation
functions (currently in major orthogonal and diagonal directions).  Such an
implementation allowed for the first time to assemble a completely free solution
to evaluate correlation functions under any operating system with well
documented application programming interface (API). Our package includes
automatic tests against analytical solutions that are described in the paper. We
measured wall times for all CPU and GPU implementations and as a rule of thumb
full correlation maps on GPU are faster than other methods. However, full maps
require more RAM and, thus, are limited to available RAM resources. On the other
hand, directional CFs are memory efficient and can be evaluated for huge
datasets -- this way they are the first candidates for structural data
compression of feature extraction. The package itself is available through Julia
package ecosystem and on GitHub
(https://github.com/fatimp/CorrelationFunctions.jl), the latter source also
contains documentation and additional helpful resources such as tutorials.  We
believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} will significantly facilitate the usage of
correlation functions in numerous areas of structural description and research,
as well as in machine learning applications. We also present some example as
applied to ceramic, soil composite and oil-bearing rock samples based on their
3D X-ray tomography and 2D scanning electron microscope images. Finally, we
conclude our paper with discussion of possible ways to further improve presented
computational framework.
\end{abstract}

@@

Correlation functions as invaluable universal descriptors of structure are
utilized in a multitude of scientific disciplines: material sciences
\cite{hasanabadi20163d,Havelka,feng2018reconstruction,chen2022quantifying}, rock
physics \cite{ledesma2018effect}, soil physics and hydrology
\cite{PLoS_ONE,KarsaninaEJSS}, cosmology and food engineering
\cite{TakadaJain,Derossi2019}, biology \cite{veatch2012correlation} and many
others. Computed from 2D and 3D images, CFs can be applied to:
\begin{enumerate}
\item characterize the morphology and representativeness via correlation
  lengths \cite{vcapek2009stochastic,thovert2011grain,tensorPRE};
\item perform stochastic reconstructions from experimentally measured CFs
  or 2D to 3D reconstructions based on CFs that can be extracted from less
  dimensions or through penetrable sphere model
  \cite{Adler_recon,Y-T,tahmasebiPRL,hasanabadi20163d,karsaninaPRL};
\item compare different structures to each other, including verification of
  stochastic reconstructions \cite{vcapek2009stochastic,tahmasebiPRL,EPL2};
\item compress structural information in the form of raw or parameterized CFs
  with the possibility to recover the structure using stochastic
  reconstruction \cite{SciRep1,Havelka,KarsaninaEJSS};
\item describe structural dynamics under different boundary conditions
  \cite{chen2015dynamic,xu2022correlation,fomin2023soil};
\item extract structural features for deep learning
  \cite{pilania2017multi,kamrava2020linking,roding2020predicting,KarsaninaEJSS};
\item fuse multi-scale images and structural information as obtained by
  different methods into a single digital model \cite{Geoderma2018};
\item quantify spatial heterogeneity \cite{zhang2000pore,REVpaper}.
\end{enumerate}

@@

For surface-surface and surface-void functions we use an edge detection filter
\cite{samarin2023robust} and then calculate either autocorrelation of the edge
or cross-correlation of the edge and the void phase. The edge is extracted by
convolving an input with a short high-pass filter $H$. It has a width $7$ and
all its coefficients with exception of the central coefficient are inversely
proportional to distance from the center:
\begin{equation}
  \begin{aligned}
    H_k &= S \left\{
    \begin{array}{ll}
      -\sum\limits_{\substack{l \\ l \ne 0}} H_l & \quad k = 0 \\
      1 / \rho(k, 0) & \quad \text{otherwise}
    \end{array}
    \right. \\
    k &\in \{-3, -2, \dots, 3\}^D
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $S$ equals to $30.45849$ in 2D and $172.96232$ in 3D case. The parameters
of the filter were adjusted based on comparison against exact computations of
surface functions on smooth-boundary sets \cite{postcov23}.

@@

The interface between phases must not change rapidly for the edge filter to
work correctly. Therefore, the algorithms for surface functions require input
images to be with an apropriate resolution. A criterion for goodness can be expressed with
$C_{\alpha}$ parameter. Let $\alpha \in [0, 1]$ and $f(x)$ be
(band-limited) input image ($x \in \mathbb{R}^2$ or $\mathbb{R}^3$). We define
$C_\alpha$ as follows:
\begin{align*}
  f_0(x) &= f(x) - \langle f(x) \rangle \\
  C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
    dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
\end{align*}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. We have previously
shown that criterion $C_{0.5} > 0.97$ holds then the image is suitable for
robust $F_{ss}$ and $F_{sv}$ evaluation \cite{samarin2023robust}.  Note that
function \code{lowfreq\_energy\_ratio} that computes $C_{0.5}$ parameter for any
input binary image is the part of \code{CorrelationFunctions.jl} package.

@@

All measurements of execution time are presented on \cref{fig:timings}. It is
immediately apparent that GPU versions (where available) are much faster than
CPU versions and limited only by amount of video memory. Functions in module
\code{Directional} (which utilize per-slice directional approach) are the
slowest among other implementations, but require less memory.

@@

\section{Summary}
\label{sec:summary}
In this paper we introduced \code{CorrelationFunctions.jl} open-source package
developed in Julia and capable of computing all classical correlation functions
based on 2D and 3D input images. Our code is capable of evaluating two-point
probability, cluster, lineal-path, surface-surface, surface-void, pore-size and
chord-length functions on both CPU and GPU architectures. There possible, we
presented two types of computations: full correlation map (correlations of each
point with other points on the image, that also allows obtaining ensemble
averaged CF) and directional correlation functions (currently in major
orthogonal and diagonal directions). Such an implementation allowed for the
first time to assemble a completely free solution to evaluate correlation
functions under any operating system with well documented API. Our package
includes automatic tests against analytical solutions that are described in this
paper. We measured wall times for all CPU and GPU implementations and as a rule
of thumb full correlation maps are faster than directional scans because
calculation of correlation maps applies fast Fourier transform to the whole
input array while, when calculating scans, FFT needs to be applied to each scan
separately. Calculation on GPU shows performance increase when compared to CPU,
as expected.  However, full maps require more RAM and, thus, are limited to
available RAM resources. On the other hand, directional CFs are memory efficient
and can be evaluated for huge datasets -- this way they are the first candidates
for structural data compression of feature extraction. The package itself is
available through Julia package ecosystem and on GitHub
(\url{https://github.com/fatimp/CorrelationFunctions.jl}), the latter source also
contains documentation and additional helpful resources such as tutorials.  We
believe that a single powerful computational tool such as
\code{CorrelationFunctions.jl} will significantly facilitate the usage of
correlation functions in numerous areas of structural description and research,
as well as in machine learning applications.
